Splitting of the homology of the punctured mapping class group

### 2019-02-18 Introduction.tex
Implemented some corrections by cfb, created gitignore file

Forgetting the punctures gives a surjective map $\ggm\to\gg$ with kernel $\bms$,
the $m-$th \emph{braid group} of the surface $\sg$: we obtain the Birman exact sequence (see \cite{Birman:mcgbr}):
\begin{equation}
\label{eq:Birman}
1\to\bms\to\ggm\to\gg\to 1.
\end{equation}
% This short exact sequence corresponds to 
% \begin{equation}
%  \cms\to\mgm\to\mg.
% \end{equation}
% Here $\cms$ is the \emph{unordered} configuration space of $m$ points in the interior of $\Sigma$, i.e.
% \begin{equation}
%  \cms=\fms/\S_m=\set{(p_1,\dots,p_m)\in\pa{\mathring\sg}^m|p_i\neq p_j \;\forall i\neq j}/\S_m,
% \end{equation}
% the quotient of the \emph{ordered} configuration space of $m$ points in the interior of $\Sigma$ by the natural
% action of the symmetric group $\S_m$ that permutes the (ordered list of) points in the configuration; $\mg$ and $\mgm$
% are the moduli spaces of Riemann structures on $\sg$ and $\sgm$ respectively.

The associated Leray-Serre spectral sequence $E(m)$ in $\Z_2-$homology has a second page $E(m)^2_{k,q}=H_k(\gg;H_q(\bms;\Z_2))$,
and converges to $H_{k+q}(\ggm;\Z_2)$.

The main theorem of this article states that this spectral sequence has already collapsed in its second page.
\begin{thm}
\label{thm:main}
For all $l\geq 0$ there is an isomorphism
\begin{equation}
\label{eq:main}
H_l(\ggm;\Z_2)\cong \bigoplus_{k+q=l} H_k(\gg;H_q(\bms;\Z_2)).
\end{equation}
\end{thm}
Thus the computation of $H_*(\ggm;\Z_2)$ reduces to the computation of the homology of $\gg$ with
twisted coefficients in the representation $H_*(\bms;\Z_2)$. We will see that the $\ggm-$representation
$H_*(\bms;\Z_2)$ splits as a direct sum of symmetric powers of $H_1(\sg)$ with symplectic action
(see theorem \ref{thm:Hbms*as*ggrep} in section \ref{sec:HBraidSurf}).

@@

The group $\gg$ is significantly smaller than $\ggm$, so the
computation of the right-hand side of equation \ref{eq:main}
may be more tractable.
%WE DO THIS COMPUTATION FOR $g=1$.

>>>>

Forgetting the punctures gives a surjective map $\ggm\to\gg$ with kernel $\bms$,
the $m$-th \emph{braid group} of the surface $\sg$. We obtain the Birman exact sequence (see \cite{Birman:mcgbr})
\begin{equation}
\label{eq:Birman}
1\to\bms\to\ggm\to\gg\to 1.
\end{equation}
% This short exact sequence corresponds to 
% \begin{equation}
%  \cms\to\mgm\to\mg.
% \end{equation}
% Here $\cms$ is the \emph{unordered} configuration space of $m$ points in the interior of $\Sigma$, i.e.
% \begin{equation}
%  \cms=\fms/\S_m=\set{(p_1,\dots,p_m)\in\pa{\mathring\sg}^m|p_i\neq p_j \;\forall i\neq j}/\S_m,
% \end{equation}
% the quotient of the \emph{ordered} configuration space of $m$ points in the interior of $\Sigma$ by the natural
% action of the symmetric group $\S_m$ that permutes the (ordered list of) points in the configuration; $\mg$ and $\mgm$
% are the moduli spaces of Riemann structures on $\sg$ and $\sgm$ respectively.

The associated Leray-Serre spectral sequence $E(m)$ in $\Z_2$-homology has a second page $E(m)^2_{k,q}=H_k(\gg;H_q(\bms;\Z_2))$,
and converges to $H_{k+q}(\ggm;\Z_2)$.

The main result of this article is that this spectral sequence collapses in its second page.
\begin{thm}
\label{thm:main}
For all $l\geq 0$ there is an isomorphism
\begin{equation}
\label{eq:main}
H_l\pa{\ggm;\Z_2}\cong \bigoplus_{k+q=l} H_k\pa{\gg;H_q\pa{\bms;\Z_2}}.
\end{equation}
\end{thm}
Thus the computation of $H_*\pa{\ggm;\Z_2}$ reduces to the computation of the homology of $\gg$ with
twisted coefficients in the representation $H_*\pa{\bms;\Z_2}$. We will see that this $\ggm$-representation
splits as a direct sum of symmetric powers of $H_1(\sg)$ with the symplectic action
(see Theorem \ref{thm:Hbms*as*ggrep}).
% in section \ref{sec:HBraidSurf}).

@@

% The group $\gg$ is significantly smaller than $\ggm$, so the
% computation of the right-hand side of equation \eqref{eq:main}
% may be more tractable.
%WE DO THIS COMPUTATION FOR $g=1$.

The strategy of the proof does not generalize to fields of characteristic different from 2 or to the \emph{pure}
mapping class group, in which we consider only diffeomorphisms of $\sgm$ that fix all punctures. In section \ref{sec:rational}
we describe in detail a counterexample with coefficients in $\Q$, which can be generalized both to coefficients
in a field $\F_p$ of odd characteristic and to the pure mapping class group.

### 2019-02-18 Preliminaries.tex
Implemented some corrections by cfb, created gitignore file

In this section we recollect some classical definitions and results about braid groups and mapping class groups.
Let $\sg$ be a compact, smooth, orientable surface of genus $g$ with one parametrised
boundary component $\partial \sg$. Let $\sgm$ be $\sg$ with a choice of $m$ distinct
points in the interior: these points are called \emph{punctures} and are not assumed to be ordered.

@@

 Notice that we require the points of the configuration to lie in the interior of $\sg$;
 $F_m(\sg)$ is a smooth, orientable $2m-$dimensional manifold.

 The symmetric group $\mathfrak{S}_m$ acts freely on $F_m(\sg)$ by permuting the points of a configuration;
 the orbit space
 \[
 F_m(\sg)/\mathfrak{S}_m
 \]
 is called the \emph{$m-$th unordered configuration space}
 of $\sg$ and is denoted by $C_m(\sg)$; this space is also a $2m-$dimensional orientable manifold
 (see figure \ref{fig:unordered}).

%  
%  We denote by $C_k(\sg)^c$ the one-point compactification of the space $C_k(\sg)$; the point
%  at infinity is denoted by $*$ and serves as basepoint.
\end{defn}

\begin{figure}\centering
 \includegraphics[scale=0.8]{figures/unordered.png}
 \caption{A configuration in the space $C_5(\Sigma_{1,1})$}
\label{fig:unordered}
\end{figure}

@@

\begin{defn}
 \label{def:mcg}
 Let $\Diff(\sg;\partial\sg)$ be the group of diffeomorphisms of $\sg$ that fix $\partial\sg$ pointwise,
 endowed with the Whitney $C^{\infty}-$topology. We denote by $\gg=\pi_0(\Diff(\sg;\partial\sg))$ its group of connected
 components, which is called the \emph{mapping class group} of $\sg$.

 Similarly let $\Diff(\sgm;\partial\sgm)$ be the group of diffeomorphisms of $\sgm$ that fix $\partial(\sg)$
 pointwise and restrict to a permutation of the $m$ punctures, and let $\ggm=\pi_0(\Diff(\sgm;\partial\sgm))$ be its
 group of connected components, called the mapping class group of $\sgm$.
 \end{defn}
 A classical result by Earle and Schatz (\cite{EarleSchatz}) ensures that the connected components
 of $\Diff(\sg)$ are contractible. In particular $B\Diff(\sg;\partial\sg)\simeq B\gg$ is
 a classifying space for
 the group $\gg$, i.e. a space of type $K(\gg,1)$; we call $\cF_{g,1}\to B\Diff(\sg;\partial\sg)$ the universal
 $\sg-$bundle
 \[
  \cF_{g,1}\colon=\sg\times_{\Diff(\sg;\partial\sg)}E\Diff(\sg;\partial\sg)\to B\Diff(\sg;\partial\sg).
 \]
Applying the construction of the $m-$th unordered configuration space fiberwise we obtain a bundle
$C_m(\cF_{g,1})\to B\Diff(\sg;\partial\sg)$ with fiber $C_m(\sg)$.

The space $C_m(\cF_{g,1})$ is a classifying space for the
group $\ggm$ and the Birman exact sequence \ref{eq:Birman} is obtained by taking fundamental
groups of the aspherical spaces
\begin{equation}\label{eq:Birmanbundle}
C_m(\sg)\to C_m(\cF_{g,1})\to B\Diff(\sg;\partial\sg).
\end{equation}

In the whole article the genus $g$ of the surfaces that we consider is supposed to be fixed,
unless explicitly stated otherwise, and we will abbreviate $\S=\sg$.
We denote by $\D$ the open disc $\mathring{\Sigma}_{0,1}$.

It will be useful, for many constructions, to choose an embedding
$\D\hookrightarrow\mrS$ near $\partial\S$ and to replace
$\Diff(\S;\partial\S)$ with its subgroup $\Diff(\S;\partial\S\cup\D)$ of diffeomorphisms
of $\S$ that fix pointwise both $\partial\S$ and $\D$. Saying that $\D$ is embedded
\emph{near $\partial\S$} means that there is a compact subsurface $\S'\subset\S$
such that $\S$ is the boundary connected sum $\S'\natural\bar\D$.
From now on we suppose such an embedding to be fixed and we consider $\D$ as a subspace
of $\mrS$ (see figure \ref{fig:SS'D}). In section \ref{sec:HBraidSurf}, definition \ref{defn:Tsg},
we will specify an embedding $\D\hookrightarrow\mrS$ using a particular model for $\mrS$.

\begin{figure}\centering
 \includegraphics[scale=0.8]{figures/SS'D.png}
 \caption{A splitting of $\S$ as $\S'\natural\bar\D$.}
\label{fig:SS'D}
\end{figure}

The construction of definition \ref{defn:cms} can be specialised to the
surfaces $\D$ and $\S'$, yielding spaces $C_m(\D)$ and $C_m(\S')$ respectively.

The inclusion $\Diff(\S;\partial\S\cup\D)\subset\Diff(\S;\partial\S)$ is a homotopy
equivalence, hence also the induced map
\[
B\Diff(\S;\partial\S\cup\D)\to B\Diff(\S;\partial\S)
\]
is a homotopy equivalence. We will replace the previous construction with the following,
equivalent ones.
\begin{defn}
\label{defn:universalSbundle}
We call $\cF_{\S,\D}\to B\Diff(\S;\partial\S\cup\D)$
the universal $\S-$bundle
 \[
  \cF_{\S,\D}\colon=\S\times_{\Diff(\S;\partial\S\cup\D)}E\Diff(\S;\partial\S\cup\D)\to B\Diff(\S;\partial\S\cup\D).
 \]
Notice that $\Diff(\S;\partial\S\cup\D)$ acts on $\S'$ by restriction of diffeomorphisms, and acts trivially
on $\D$; therefore $\cF_{\S,\D}$ contains subspaces
\[
 \cF_{\S'}= \S'\times_{\Diff(\S;\partial\S\cup\D)}E\Diff(\S;\partial\S\cup\D)
\]
and $\D\times B\Diff(\S;\partial\S\cup\D)$; these subspaces fiber over $B\Diff(\S;\partial\S\cup\D)$
with fibers $\S'$ and $\D$ respectively.

Applying the construction of the $m-$th unordered configuration space fiberwise, we obtain spaces
$C_m(\cF_{\S,\D})$, $C_m(\cF_{\S'})$ and $C_m(\D)\times B\Diff(\S;\partial\S\cup\D)$,
all fibering over $B\Diff(\S;\partial\S\cup\D)$ with fibers, respectively, $\cms$,
$C_m(\S')$ and $C_m(\D)$.
\end{defn}

\begin{defn}
For all $0\leq p\leq m$ there is a natural map
\[
 \mu\colon C_p(\D)\times C_{m-p}(\S')\to \cms
\]
which takes the union of configurations (see figure \ref{fig:defmu}):
 \[
  \mu\pa{\set{P_1,\dots,P_p};\set{P'_1,\dots,P'_{m-p}}}=\set{P_1,\dots,P_p,P'_1,\dots,P'_{m-p}}\in\cms.
 \]
\begin{figure}\centering
 \includegraphics[scale=0.7]{figures/defmu.png}
 \caption{The product of two configurations in $C_2(\D)$ and $C_4(\S)$.}
\label{fig:defmu}
\end{figure}

We can apply this construction to each couple of fibers of $C_m(\cF_{\S,\D})$ and $C_{m-p}(\cF_{\S'})$
over the same point of $ B\Diff(\S;\partial\S\cup\D)$, obtaining a map $\mu^{\cF}$:
\[
 \mu^{\cF}\colon C_p(\D)\times C_{m-p}(\cF_{\S'})\to C_m(\cF_{\S,\D}).
\]
If we see the domain of $\mu^{\cF}$ as a fibered product of bundles over $B\Diff(\S;\partial\S\cup\D)$
\[
 \pa{C_m(\D)\times B\Diff(\S;\partial\S\cup\D)}\times_{B\Diff(\S;\partial\S\cup\D)}C_{m-p}(\cF_{\S'}),
\]
then the map $\mu^{\cF}$ is also a map of bundles over $B\Diff(\S;\partial\S\cup\D)$, and the corresponding
map on fibers is precisely $\mu$.
\end{defn}
The fiber bundle \ref{eq:Birmanbundle} corresponding to the Birman exact sequence \ref{eq:Birman}
can now be replaced by the following one
\begin{equation}\label{eq:BirmanbundleD}
\cms\to C_m(\cF_{\S,\D})\to B\Diff(\S;\partial\S\cup\D).
\end{equation}
We conclude this section by recalling the structure of $H_*(C_m(\D))$.

The cohomology of $C_m(\D)$ with coefficients in $\Z_2$ was first computed by Fuchs in
\cite{Fuchs:CohomBraidModtwo}: we will recall this computation in section
\ref{sec:HBraidSurf}, where we will make a similar computation in detail.

In \cite[Chap.III]{CLM} Cohen considers the space $\coprod_{m\geq 0}C_m(\D)$ as an
algebra over the operad of little $2-$cubes, and describes its
$\Z_2-$homology as follows:
\begin{equation}
 \label{eq:Cohen}
H_*\pa{\coprod_{m\geq 0}C_m(\D)}\simeq \Z_2\left[Q^j\epsilon \, |\, j\geq 0\right].
\end{equation}

@@

The isomorphism in equation \ref{eq:Cohen} is
an isomorphism of bigraded rings. The left-hand side is a ring with the Pontryagin product
coming from the action of the little $2-$cubes,
and the right-hand side is a polynomial ring in infinitely many variables $\epsilon,Q\epsilon,Q^2\epsilon,\dots$.
The bidegree is given by the homological degree
$*$, that we simply call the \emph{degree},
and by the index $m$ of the connected component
on which the homology class is supported (informally, the number of points
involved in the construction of a homology class), that we call the \emph{weight}.

 \begin{figure}\centering
 \includegraphics[scale=0.7]{figures/monomial.png}
 \caption{The class in $H_5(C_9(\D))$ corresponding to the monomial $\epsilon\cdot(Q\epsilon)^2\cdot Q^2\epsilon$.}
\label{fig:monomial}
\end{figure}

In this article we will only need the isomorphism in equation \ref{eq:Cohen} to hold as
an isomorphism of bigraded $\Z_2-$vector spaces.
In particular for all choices of natural numbers $(\alpha_j)_{j\geq 0}$ with all but finitely
many $\alpha_j=0$, we will consider the monomial $\prod_j\pa{Q^j\epsilon}^{\alpha_j}$, corresponding to
a homology class in some bidegree $(*,m)$: all that
we are going to need is that the set of these monomials is a bigraded basis for the
left-hand side of equation \ref{eq:Cohen}.

>>>>

In this section we recollect some classical definitions and results about braid groups and mapping class groups.
% Let $\sg$ be a compact, smooth, orientable surface of genus $g$ with one parametrised
% boundary component $\partial \sg$. Let $\sgm$ be $\sg$ with a choice of $m$ distinct
% points in the interior: these points are called \emph{punctures} and are not assumed to be ordered.

@@

 Note that we require the points of the configuration to lie in the interior of $\sg$; the space
 $F_m(\sg)$ is a smooth, orientable $2m$-dimensional manifold.

 The symmetric group $\mathfrak{S}_m$ acts freely on $F_m(\sg)$ by permuting the labels $1,\dots,n$ of a configuration;
 the orbit space
 \[
 F_m(\sg)/\mathfrak{S}_m
 \]
 is called the \emph{$m$-th unordered configuration space}
 of $\sg$ and is denoted by $C_m(\sg)$; this space is also a $2m$-dimensional orientable manifold
 (see figure \ref{fig:unordered}).

%  
%  We denote by $C_k(\sg)^{\infty}$ the one-point compactification of the space $C_k(\sg)$; the point
%  at infinity is denoted by $*$ and serves as basepoint.
\end{defn}

\begin{figure}\centering
\includegraphics[scale=0.8]{figures/unordered.pdf}
\caption{A configuration in the space $C_5(\Sigma_{1,1})$}
\label{fig:unordered}
\end{figure}

@@

\begin{defn}
 \label{def:mcg}
 Let $\Diff(\sg;\partial\sg)$ be the group of diffeomorphisms of $\sg$ that fix $\partial\sg$ pointwise,
 endowed with the Whitney $C^{\infty}$-topology. We denote by $\gg=\pi_0(\Diff(\sg;\partial\sg))$ its group of connected
 components, which is called the \emph{mapping class group} of $\sg$.

 Similarly let $\Diff(\sgm;\partial\sgm)$ be the group of diffeomorphisms of $\sgm$ that fix $\partial(\sg)$
 pointwise and restrict to a permutation of the $m$ punctures: the \emph{mapping class group} of $\sgm$ is
 $\ggm=\pi_0(\Diff(\sgm;\partial\sgm))$.
 \end{defn}
 A classical result by Earle and Schatz \cite{EarleSchatz} ensures that the connected components
 of $\Diff(\sg)$ are contractible. In particular $B\Diff(\sg;\partial\sg)\simeq B\gg$ is
 a classifying space for
 the group $\gg$, i.e. an Eilenberg-MacLane space of type $K(\gg,1)$.

 We call $\cF_{g,1}\to B\Diff(\sg;\partial\sg)$ the universal
 $\sg$-bundle
 \[
  \sg\to\cF_{g,1}=\sg\times_{\Diff(\sg;\partial\sg)}E\Diff(\sg;\partial\sg)\to B\Diff(\sg;\partial\sg).
 \]
Applying the construction of the $m$-th unordered configuration space fiberwise we obtain a bundle
$C_m(\cF_{g,1})\to B\Diff(\sg;\partial\sg)$ with fiber $C_m(\sg)$.

The space $C_m(\cF_{g,1})$ is a classifying space for the
group $\ggm$ and the Birman exact sequence \eqref{eq:Birman} is obtained by taking fundamental
groups of the aspherical spaces
\begin{equation}\label{eq:Birmanbundle}
C_m(\sg)\to C_m(\cF_{g,1})\to B\Diff(\sg;\partial\sg).
\end{equation}

In the whole article the genus $g\geq 0$ of the surfaces that we consider is supposed to be fixed,
% unless explicitly stated otherwise,
and we will abbreviate $\S=\sg$.
We denote by $\D$ the open disc $\mathring{\Sigma}_{0,1}$.

It will be useful, for many constructions, to choose an embedding
$\D\hookrightarrow\mrS$ \emph{near} $\partial\S$ and to replace
$\Diff(\S;\partial\S)$ with its subgroup $\Diff(\S;\partial\S\cup\D)$ of diffeomorphisms
of $\S$ that fix pointwise both $\partial\S$ and $\D$. Saying that $\D$ is embedded
\emph{near $\partial\S$} means that there is a compact subsurface $\S'\subset\S$
such that $\S=\S'\natural\bar\D$ is the union along a segment of $\S'$ and the closure of $\D$ in $\S$.

From now on we suppose such an embedding to be fixed and we consider $\D$ as a subspace
of $\mrS$ (see Figure \ref{fig:SS'D}). In Section \ref{sec:HBraidSurf}, Definition \ref{defn:Tsg},
we will introduce a convenient model $\T(\S)$ for the space $\mrS$, and in Section \ref{sec:Actiongg},
Definition \ref{defn:DinTS}
we will specify an embedding $\D\hookrightarrow\mrS$ using the model $\T(\S)$.

\begin{figure}\centering
 \includegraphics[scale=0.8]{figures/SS'D.pdf}
 \caption{A splitting of $\S$ as $\S'\natural\bar\D$.}
\label{fig:SS'D}
\end{figure}

The construction in Definition \ref{defn:cms} can be specialised to the
surfaces $\D$ and $\S'$, yielding spaces $C_m(\D)$ and $C_m(\S')$ respectively.
We take configurations of points in the interior of the surfaces $\D$ and $\S'$ respectively
(since $\D$ is already an open surface this remark applies in particular to $\S'$).

The inclusion $\Diff(\S;\partial\S\cup\D)\subset\Diff(\S;\partial\S)$ is a homotopy
equivalence of topological groups, hence also the induced map
\[
B\Diff(\S;\partial\S\cup\D)\to B\Diff(\S;\partial\S)
\]
is a homotopy equivalence. We will replace the previous construction with the following,
homotopy equivalent ones.
\begin{defn}
\label{defn:universalSbundle}
We call $\cF_{\S,\D}\to B\Diff(\S;\partial\S\cup\D)$
the universal $\S$-bundle
 \[
  \S\to\cF_{\S,\D}\colon=\S\times_{\Diff(\S;\partial\S\cup\D)}E\Diff(\S;\partial\S\cup\D)\to B\Diff(\S;\partial\S\cup\D).
 \]
Note that $\Diff(\S;\partial\S\cup\D)$ acts on $\S'$ by restriction of diffeomorphisms, and acts trivially
on $\D$; therefore $\cF_{\S,\D}$ contains subspaces
\[
 \cF_{\S'}= \S'\times_{\Diff(\S;\partial\S\cup\D)}E\Diff(\S;\partial\S\cup\D)
\]
and $\D\times B\Diff(\S;\partial\S\cup\D)$; these subspaces fiber over $B\Diff(\S;\partial\S\cup\D)$
with fibers $\S'$ and $\D$ respectively.

Applying the construction of the $m$-th unordered configuration space fiberwise, we obtain spaces
$C_m(\cF_{\S,\D})$, $C_m(\cF_{\S'})$ and $C_m(\D)\times B\Diff(\S;\partial\S\cup\D)$,
all fibering over $B\Diff(\S;\partial\S\cup\D)$, respectively with fiber $\cms$,
$C_m(\S')$ and $C_m(\D)$.
\end{defn}
The fiber bundle \eqref{eq:Birmanbundle} corresponding to the Birman exact sequence \eqref{eq:Birman}
can now be replaced with the following one
\begin{equation}\label{eq:BirmanbundleD}
\cms\to C_m(\cF_{\S,\D})\to B\Diff(\S;\partial\S\cup\D).
\end{equation}

\begin{defn}
For all $0\leq p\leq m$ there is a map
\[
 \mu\colon C_p(\D)\times C_{m-p}(\S')\to \cms
\]
which takes the union of configurations (see figure \ref{fig:defmu}):
 \[
  \mu\pa{\set{P_1,\dots,P_p};\set{P'_1,\dots,P'_{m-p}}}=\set{P_1,\dots,P_p,P'_1,\dots,P'_{m-p}}\in\cms.
 \]
\begin{figure}\centering
 \includegraphics[scale=0.7]{figures/defmu.pdf}
 \caption{The product of two configurations in $C_2(\D)$ and $C_4(\S')$.}
\label{fig:defmu}
\end{figure}

We can apply this construction at the same time to each couple of fibers of the bundles $C_m(\cF_{\S,\D})$ and $C_{m-p}(\cF_{\S'})$
lying over the same point of $ B\Diff(\S;\partial\S\cup\D)$.

We obtain a map
\[
 \mu^{\cF}\colon C_p(\D)\times C_{m-p}(\cF_{\S'})\to C_m(\cF_{\S,\D}).
\]
If we see the domain of $\mu^{\cF}$ as a fibered product of bundles over $B\Diff(\S;\partial\S\cup\D)$
\[
 \pa{C_m(\D)\times B\Diff(\S;\partial\S\cup\D)}\times_{B\Diff(\S;\partial\S\cup\D)}C_{m-p}(\cF_{\S'}),
\]
then the map $\mu^{\cF}$ is also a map of bundles over $B\Diff(\S;\partial\S\cup\D)$, and the corresponding
map on fibers is precisely $\mu$.
\end{defn}

We recall now the structure of $H_*(C_m(\D))$.
The cohomology of $C_m(\D)$ with coefficients in $\Z_2$ was first computed by Fuchs in
\cite{Fuchs:CohomBraidModtwo}: in section
\ref{sec:HBraidSurf} we will generalise  Fuchs' argument to surfaces with boundary of positive genus.

In \cite[Chap.III]{CLM} Cohen considers the space $\coprod_{m\geq 0}C_m(\D)$ as an
algebra over the operad of little $2$-cubes, and describes its
$\Z_2$-homology as follows:
\begin{equation}
 \label{eq:Cohen}
H_*\pa{\coprod_{m\geq 0}C_m(\D)}\simeq \Z_2\left[Q^j\epsilon \, |\, j\geq 0\right].
\end{equation}

@@

The isomorphism in equation \eqref{eq:Cohen} is
an isomorphism of bigraded rings. The left-hand side is a ring with the Pontryagin product,
% coming from the action of the little $2$-cubes,
and the right-hand side is a polynomial ring in infinitely many variables $\epsilon,Q\epsilon,Q^2\epsilon,\dots$.
The bigrading is given by the homological degree
$*$, that we call the \emph{degree},
and by the index $m$ of the connected component
on which the homology class is supported (informally, the number of points
involved in the construction of the homology class), that we call the \emph{weight}.

 \begin{figure}\centering
 \includegraphics[scale=0.7]{figures/monomial.pdf}
 \caption{The class in $H_5(C_9(\D))$ corresponding to the monomial $\epsilon\cdot(Q\epsilon)^2\cdot Q^2\epsilon$.}
\label{fig:monomial}
\end{figure}

In this article we will only need the isomorphism in equation \eqref{eq:Cohen} to hold as
an isomorphism of bigraded $\Z_2$-vector spaces.
In particular for all choices of natural numbers $(\alpha_j)_{j\geq 0}$ with all but finitely
many $\alpha_j=0$, we will consider the monomial $\prod_j\pa{Q^j\epsilon}^{\alpha_j}$, corresponding to
a homology class in some bigrading $(*,m)$: we will only use the fact
that the set of these monomials is a bigraded basis for the
left-hand side of equation \eqref{eq:Cohen}.

We end this section recalling the classical notions of symmetric product and one-point compactification.

\begin{defn}
 \label{defn:SP}
 The $m$-fold symmetric product of a topological space $\mathcal{M}$, denoted by $\SP^m(\mathcal{M})$, is
 the quotient of $\mathcal{M}^m$ by the (non-free) action of the symmetric group $\mathfrak{S}_m$
 which permutes the coordinates.
\end{defn}
We regard a point of $\SP^m(\mathcal{M})$ as a configuration of $m$ points in $\mathcal{M}$ with multiplicities.

In the case of an open surface $\mrS$, it is known that $\SP^m(\mrS)$ is a non-compact manifold of dimension $2m$;
it contains $\cms$ as an open submanifold.

\begin{defn}
 \label{defn:opc}
For a space $\mathcal{M}$ we denote by $\mathcal{M}^{\infty}$ its one-point
compactification; the basepoint is the point at infinity, that we denote by $\infty$.
\end{defn}

We consider in particular the one-point-compactification $\cms^{\infty}$ of $\cms$.
